---
title: "Visualizing NWIS data"
author: 'Laura Erban'
date: "`r Sys.Date()`"
output:
  html_document:
    theme: yeti
    code_folding: hide
    number_sections: no
    toc: no
    toc_float: no
  html_notebook:
    theme: yeti
    code_folding: hide
    number_sections: no
    toc: no
    toc_float: no
---

```{r load_libraries, include=FALSE}
library(here)
library(leaflet)
library(leafsync)
library(leaflegend)
library(dataRetrieval)
library(plotly)
library(knitr)
library(tidyverse)
```

$~$

This notebook helps visualize and explore surface and groundwater observations stored in the National Water Information System (NWIS). It was written to perform automated pulls and interactive mapping and plotting of data as it was being collected and analyzed for a specific project, but it can be modified for any region where there is data in NWIS.  

$~$

### 1. Map sites 
```{r}
# watershed screening sites 
screensites <- whatNWISsites(bBox = c(-70.448, 41.626, -70.392, 41.677))
nums <- "504|505|506|507"
screensites <- screensites %>%
  filter(grepl(nums, station_nm))

# local demonstration sites
sites <- whatNWISsites(bBox = c(-70.405, 41.669, -70.395, 41.677))

# site info
baseurl1 <- "https://nwis.waterdata.usgs.gov/usa/nwis/qwdata/?site_no="
baseurl2 <-  "&agency_cd=USGS"

siteinfo <- readNWISsite(sites$site_no) 
#comment(siteinfo)
siteinfo <- siteinfo %>%
  dplyr::select(site_no, station_nm, site_tp_cd, dec_lat_va, dec_long_va, dec_coord_datum_cd, alt_va, alt_datum_cd, well_depth_va)%>%
  mutate(id = substr(station_nm, 9, 11)) %>%
  mutate(name = substr(station_nm,1,11)) %>%
  group_by(id) %>% mutate(depthsatsite = n()) %>% ungroup() %>% 
  mutate(weblink = paste(baseurl1,site_no,baseurl2,sep = " "))

# reclassify site types for mapping
siteinfo$site_tp_cd[grep("M01", siteinfo$station_nm)] <- "multilevel sampler (MLS)"
clust <- "505-0|517-0|524-0|525-0|526-0"
siteinfo$site_tp_cd[grepl(clust, siteinfo$station_nm)] <- "well cluster"
siteinfo$site_tp_cd[which(siteinfo$site_tp_cd=="GW")] <- "water table well"
siteinfo$site_tp_cd[which(siteinfo$site_tp_cd=="LK")] <- "surface water (pond)"

```

```{r}
# map all sites by type
pal <- colorFactor("Accent", domain = siteinfo$site_tp_cd)

leaflet(siteinfo, options = leafletOptions(zoomControl = FALSE)) %>%
  addProviderTiles('OpenStreetMap.Mapnik')%>%
  addCircleMarkers(lng = ~dec_long_va, lat = ~dec_lat_va, weight = 1, color = "gray", radius = 6, opacity = 1, 
                 fillOpacity = 1, fillColor = ~pal(site_tp_cd),
                 popup = ~paste(name,"<br>", "Depths: ", depthsatsite)) %>%
  addScaleBar("bottomright") %>%
  addLegend(pal = pal, title = "Site type", opacity = 1, values = ~site_tp_cd, position = "topleft")

```

$~$

### 2. Specify water quality parameters for data retrieval
``` {r}
# parameters     (nutrients samples are all filtered)
temp <- "00010"  # temperature                 (degC)
sc <- "00095"    # SpC                         (uS/cm)
ph <- "00400"    # pH
do <- "00300"    # DO                          (mg/L)
nh34 <- "00608"  # ammonia and ammonium, as N  (mg/L)
no2 <- "00613"   # nitrite as N                (mg/L)
no3 <- "00618"   # nitrate as N                (mg/L)
no23 <- "00631"  # nitrate + nitrite as N      (mg/L)
tn <- "62854"    # total nitrogen              (mg/L)
po4 <- "00671"   # orthophosphate as P         (mg/L)
delH <- "82082"  # delta H2/H1                (per mil)
delO <- "82085"  # delta O18/O16              (per mil)

pset <- c(temp, sc, ph, do, nh34, no2, no3, no23, tn, po4, delH, delO)
```

### 3. Retrieve and summarize data

```{r retrieve, eval = TRUE, include = TRUE, message=FALSE, warning=FALSE}
# RETRIEVAL

# pull water quality data
## option A (to be retired)
data <- readNWISqw(siteNumber = siteinfo$site_no, parameterCd = pset)
b <- data %>% dplyr::select(c(site_no, sample_dt, sample_tm, parm_cd, result_va))

# # ## option B (suggested going forward)
# datab <- readWQPqw(siteNumber = paste0("USGS-", siteinfo$site_no), parameterCd = pset)
# # select columns and rename
# b <- datab %>% dplyr::select(c(MonitoringLocationIdentifier, ActivityStartDate, USGSPCode, CharacteristicName, ResultMeasure.MeasureUnitCode, ResultMeasureValue))
# names(b) <- c("site_no", "sample_dt", "parm_cd", "charname", "units", "result_va")
# b$site_no <- sub("USGS-", "", b$site_no)

# combine site info and qw data
a <- siteinfo 
a$station_nm[grep("SHUBAEL", a$station_nm)] <- "SHUBAEL POND"
alldata <- right_join(a,b) %>%
  dplyr::rename(date = sample_dt) %>%
  mutate(param = parm_cd, yearmon = format(date, "%Y-%m")) 

# create param names from codes 
alldata$param <- dplyr::recode(alldata$param, `00010` = "temp", `00095` = "SpC", `00400` = "pH", `00300` = "DO", `00608` = "NH34_N", `00613` = "NO2_N", `00618` = "NO3_N", `00631` = "NO23_N", `62854` = "TN", `00671` = "PO4_P", `82082` = "delH", `82085` = "delO")

# pull water levels data
gwl <- readNWISgwl(siteNumbers = siteinfo$site_no) 

# join site info
dtw <- gwl %>%
  filter(!is.na(lev_va))
dtw <- right_join(siteinfo %>% dplyr::select(station_nm, site_no, id, dec_lat_va, dec_long_va, alt_va, well_depth_va, weblink), dtw)%>%
  dplyr::rename(date = lev_dt) %>%
  mutate(yearmon = format(date, "%Y-%m"))

```

```{r summarize, eval = TRUE, include = TRUE, message=FALSE, warning=FALSE}
# SUMMARIES

# number of sample depths at each location
nwells <- alldata %>% 
  group_by(id) %>%
  summarise(depths = length(unique(well_depth_va))) 

# number of depths and samples at each location, by parameter
allsum <- alldata %>% 
  group_by(id, param) %>%
  dplyr::summarise(count = n()) %>%
  pivot_wider(id_cols = id, names_from = param, values_from = count)

allsum <- left_join(allsum, nwells) %>%
  dplyr::select(id, depths, temp, SpC, pH, DO, PO4_P, NH34_N:NO3_N, TN) %>%
  arrange(depths)

# subset data for a specific sampling event
ym <- "2022-03"
zt <- alldata %>%
  #dplyr::filter(yearmon == ym)
  dplyr::filter(date > "2022-03-05") %>%
  dplyr::filter(date < "2022-04-01")

# number of depths and samples at each location, by parameter
ztsum <- zt %>% 
  group_by(id, param) %>%
  summarise(count = n()) %>%
  pivot_wider(id_cols = id, names_from = param, values_from = count)

ztsum <- left_join(ztsum, nwells) %>%
  dplyr::select(id, depths, temp, SpC, pH, DO, PO4_P, NH34_N:NO3_N, TN) %>%
  arrange(depths)

# means at each location, single time
zmean <- zt %>%
  group_by(id, parm_cd) %>%
  summarise(station_nm=first(id), dec_lat_va=median(dec_lat_va), dec_long_va=median(dec_long_va), weblink=first(weblink), well_depth_va=mean(well_depth_va, na.rm=TRUE), result_va=round(mean(result_va, na.rm=TRUE), 3))

# means at each location, time series
tsmean <- alldata %>%
  group_by(id, yearmon, param) %>%
  summarise(station_nm=first(id), dec_lat_va=median(dec_lat_va), dec_long_va=median(dec_long_va), weblink=first(weblink), well_depth_va=mean(well_depth_va, na.rm=TRUE), result_va=round(mean(result_va, na.rm=TRUE), 3))

# max over sample depths at each location
zmax <- zt %>%
  group_by(id, parm_cd) %>%
  summarise(station_nm=first(id), dec_lat_va=median(dec_lat_va), dec_long_va=median(dec_long_va), weblink=first(weblink), well_depth_va=mean(well_depth_va, na.rm=TRUE), result_va=max(result_va, na.rm=TRUE))
```


```{r}

# calculate percentages of N in different forms
ftn <- pivot_wider(zt, id_cols = c(id, site_no:dec_long_va, well_depth_va), names_from = param, values_from = result_va)%>%
  filter(!is.na(TN)) %>%
  mutate(percent_organicN = round(100*(1 - ((NH34_N+NO23_N)/TN)), 1), percent_NO23_N = round(100*NO23_N/TN, 1))

kable(ftn %>% arrange(id) %>% dplyr::select(station_nm, temp, SpC, pH, DO, PO4_P, NH34_N, NO3_N, NO23_N, TN, percent_organicN, percent_NO23_N), digits = 3, col.names = c('Station', 'Temp (ºC)', 'SpC (µS/cm)', 'pH (-)', 'DO', 'PO4-P', 'NH34-N', 'NO3-N', 'NO23-N', 'TN', 'org. N (%)', 'NO23-N (%)'), caption = paste0("Snapshot of ", ym, " sampling results. All units are mg/L unless otherwise noted."))

```

$~$

$~$

### 4. Map water quality parameters and depth to water

$~$

```{r}

# function to make the maps
mapparam <- function(data, palette, title){
  leaflet(data, options = leafletOptions(zoomControl = TRUE, attributionControl=FALSE)) %>%
  addProviderTiles('OpenStreetMap.Mapnik')%>%
  addCircleMarkers(lng = ~dec_long_va, lat = ~dec_lat_va, weight = 1, color = "gray", radius = 6, opacity = 1, 
                   fillOpacity = 1, fillColor = ~pal(result_va),
                   popup = ~paste(station_nm,"<br>","Value",result_va)) %>%  # ,"<br>","<a href='", weblink, "' target='_blank'>", "Link to data</a>")) %>%
  addScaleBar("bottomright") %>%
  addLegend(pal = palette, title = title, opacity = 1, values = ~result_va, position = "bottomleft")
}

```


### 4a. Mean values at each location for depths sampled in `r ym`


```{r, eval = TRUE, include = TRUE, message=FALSE, warning=FALSE}
z <- zmean

ttl <- "Temp (°C)"
mapdata <- z %>% filter(parm_cd == temp)
pal <- colorBin("viridis", domain = mapdata$result_va, bins = 5)
m1 <- mapparam(mapdata, pal, ttl)

ttl <- "SpC (uS/cm)"
mapdata <- z %>% filter(parm_cd == sc)
pal <- colorBin("viridis", domain = mapdata$result_va, bins = 5)
m2 <- mapparam(mapdata, pal, ttl)

ttl = "pH"
mapdata <- z %>% filter(parm_cd == ph)
pal <- colorBin("viridis", domain = mapdata$result_va, bins = 5)
m3 <- mapparam(mapdata, pal, ttl)

ttl = "DO (mg/L)"
mapdata <- z %>% filter(parm_cd == do)
pal <- colorBin("viridis", domain = mapdata$result_va, bins = 5)
m4 <- mapparam(mapdata, pal, ttl)

ttl = "NO3, as N <br> (mg/L)"
mapdata <- z %>% filter(parm_cd == no3)
pal <- colorBin("viridis", domain = mapdata$result_va, bins = 5)     
m5 <- mapparam(mapdata, pal, ttl)

ttl = "NH34, as N <br> (mg/L)"
mapdata <- z %>% filter(parm_cd == nh34)
pal <- colorBin("viridis", domain = mapdata$result_va, bins = 5)
m6 <- mapparam(mapdata, pal, ttl)

ttl = "TN (mg/L)"
mapdata <- z %>% filter(parm_cd == tn)
pal <- colorBin("viridis",  domain = mapdata$result_va, bins = 5)   
m7 <- mapparam(mapdata, pal, ttl)

ttl = "PO4, as P <br> (mg/L)"
mapdata <- z %>% filter(parm_cd == po4)
pal <- colorBin("viridis", domain = mapdata$result_va, bins = 5)    
m8 <- mapparam(mapdata, pal, ttl)

# map depth to water
pal <- colorBin("viridis", domain = dtw$lev_va, bins = 5)
m9 <- leaflet(dtw %>% filter(yearmon == ym), options = leafletOptions(zoomControl = FALSE, attributionControl=FALSE)) %>%
  addProviderTiles('OpenStreetMap.Mapnik')%>%
  addCircleMarkers(lng = ~dec_long_va, lat = ~dec_lat_va, weight = 1, color = "gray", radius = 6, opacity = 1, 
                 fillOpacity = 1, fillColor = ~pal(lev_va),
                 popup = ~paste(station_nm,"<br>","Value",lev_va,"<br>","<a href='", weblink, "' target='_blank'>", "Link to data</a>")) %>%
  addScaleBar("bottomright") %>%
  addLegend(pal = pal, title = "Depth to water", opacity = 1, values = ~lev_va, position = "bottomleft")

sync(m1, m2, m3, m4, m5, m6, m7, m8, m9, ncol = 3, sync.cursor = FALSE)
```

```{r} 
# this is a dummy chunk to separate maps from next section
```

### 4b. Maximum values at each location for depths sampled in `r ym`

$~$

```{r, eval = TRUE, include = TRUE, message=FALSE, warning=FALSE}
z <- zmax

ttl <- "Temp (°C)"
mapdata <- z %>% filter(parm_cd == temp)
pal <- colorBin("viridis", domain = mapdata$result_va, bins = 5)
m1 <- mapparam(mapdata, pal, ttl)

ttl <- "SpC (uS/cm)"
mapdata <- z %>% filter(parm_cd == sc)
pal <- colorBin("viridis", domain = mapdata$result_va, bins = 5)
m2 <- mapparam(mapdata, pal, ttl)

ttl = "pH"
mapdata <- z %>% filter(parm_cd == ph)
pal <- colorBin("viridis", domain = mapdata$result_va, bins = 5)
m3 <- mapparam(mapdata, pal, ttl)

ttl = "DO (mg/L)"
mapdata <- z %>% filter(parm_cd == do)
pal <- colorBin("viridis", domain = mapdata$result_va, bins = 5)
m4 <- mapparam(mapdata, pal, ttl)

ttl = "NO3, as N <br> (mg/L)"
mapdata <- z %>% filter(parm_cd == no3)
pal <- colorBin("viridis", domain = mapdata$result_va, bins = 5)     
m5 <- mapparam(mapdata, pal, ttl)

ttl = "NH34, as N <br> (mg/L)"
mapdata <- z %>% filter(parm_cd == nh34)
pal <- colorBin("viridis", domain = mapdata$result_va, bins = 5)
m6 <- mapparam(mapdata, pal, ttl)

ttl = "TN (mg/L)"
mapdata <- z %>% filter(parm_cd == tn)
pal <- colorBin("viridis",  domain = mapdata$result_va, bins = 5)   
m7 <- mapparam(mapdata, pal, ttl)

ttl = "PO4, as P <br> (mg/L)"
mapdata <- z %>% filter(parm_cd == po4)
pal <- colorBin("viridis", domain = mapdata$result_va, bins = 5)    
m8 <- mapparam(mapdata, pal, ttl)

sync(m1, m2, m3, m4, m5, m6, m7, m8, m9, ncol = 3, sync.cursor = FALSE)
```


$~$


```{r}
b = 0   
```


### 5. Plot time series 

$~$

```{r, eval = TRUE, include = TRUE}
ttl = "Nitrate, as N, (mg/L) in groundwater wells and pond"
p <- ggplot(alldata %>% filter(parm_cd==no3), aes(x = date, y = result_va, group = station_nm)) +
  geom_point(aes(color=id)) +
  geom_line(aes(color=id)) +
  theme_bw()+ 
  ggtitle(ttl) 

ggplotly(p)

```

$~$

### 6. Plot vertical profiles

$~$

```{r, eval = TRUE, include = TRUE, message = FALSE}
ttl = "NO3-N (mg/L) in vertical profiles"

a <- alldata %>% filter(parm_cd==no3) %>% 
  filter(id %in% c(505, 508, 517, 522, 523, 524, 525, 526)) %>% 
  filter(station_nm != "MA-A1W  517-0036") %>% # duplicate well @ 517 
  mutate(altitude = alt_va - well_depth_va)

p <- ggplot(a, aes(x = result_va, y = altitude)) +
  geom_path(aes(color = yearmon))+
  facet_wrap(~id, ncol = 4)+                          # scales = "free",
  coord_cartesian(ylim = c(15, 45)) +
  theme(panel.spacing.x = unit(0.5, "cm"), panel.spacing.y = unit(1, "cm"), axis.text.y = element_text(vjust = -5)) +
  #theme_bw() + 
  #scale_y_reverse()+ #limits = yrange
  ggtitle(ttl)

ggplotly(p)

ttl = "PO4-P (mg/L) in vertical profiles"

a <- alldata %>% filter(parm_cd==po4) %>% 
  filter(id %in% c(505, 508, 517, 522, 523, 524, 525, 526)) %>% 
  filter(station_nm != "MA-A1W  517-0036") %>% # duplicate well @ 517 
  mutate(altitude = alt_va - well_depth_va)

p <- ggplot(a, aes(x = result_va, y = altitude)) +
  geom_path(aes(color = yearmon))+
  facet_wrap(~id, ncol = 4)+                          # scales = "free",
  coord_cartesian(ylim = c(15, 45)) +
  theme(panel.spacing.x = unit(0.5, "cm"), panel.spacing.y = unit(1, "cm"), axis.text.y = element_text(vjust = -5)) +
  #theme_bw() + 
  #scale_y_reverse()+ #limits = yrange
  ggtitle(ttl)

ggplotly(p)
```





